---
title: "Google Data Analytics Certificate Case Study"
subtitle: "Cyclystic - How does a bike-share navigate speedy success?"
author: "Austin Seo"
date: "12/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
The following will contain my case study on the fictional bike-share company, Cyclystic. This case study is in relation to the Google Data Analytics Professional Certificate. I received the data from this program through this link. However, the data has been made available Motivate International Inc. under this license. 


### About the Company
In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

Until now, Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments.One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members. 

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, Moreno believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers, Moreno believes there is a very good chance to convert casual riders into members. She notes that casual riders are already aware of the Cyclistic program and have chosen Cyclistic for their mobility needs.


### Business Problem
Moreno has set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would buy a membership, and how digital media could affect their marketing tactics. Moreno and her team are interested in analyzing the Cyclistic historical bike trip data to identify trends.

#### The goal of this case study is to answer the question: "How do annual members and casual riders use Cyclistic bikes differently?"  

## 1 - Load Packages
```{r load_packages, message=FALSE}
library(ggplot2)
library(tidyverse)
library(here)
library(janitor)
library(skimr)
library(readr)
library(plyr)
library(lubridate)
library(maptools)
library(mapproj)
library(mapdata)
library(ggmap)
```

## 2 - Load Data
```{r load_data, message=FALSE, cache=TRUE}
all_trip_data <- list.files(path = "./CSV", pattern = "*.csv", full.names = TRUE) %>% 
  lapply(read_csv, na.omit) %>% 
  bind_rows
```

## 3 - Clean Data
In the following sections, I will detail my process of cleaning the data.

### 3.1 - Create Data Frame of data
```{r}
all_trips_df <- data.frame(all_trip_data)
head(all_trips_df)
```
```{r}
colnames(all_trips_df)
```


### 3.2 - Rename Columns

Looks like we're going to have to change the column names before we proceed.

```{r}
names(all_trips_df) [1:13] <- c("ride_id", "rideable_type", "started_at", "ended_at", "start_station_name", 
                                "start_station_id", "end_station_name", "end_station_id", "start_lat", "start_lng", 
                                "end_lat", "end_lng", "member_casual")
```

We will need to remove the first row of the df as it contains what was supposed to be the column names
```{r}
all_trips_df <- all_trips_df[-c(1),]
```

Let's check out what our df looks like now...
```{r}
head(all_trips_df)
```

### 3.3 - Drop NAs
Before proceeding, I dropped all rows that contained an NA value. 
```{r}
all_trips_df <- all_trips_df %>% 
  drop_na() # drops rows with NA
```


Let's take a look at what we have now.
```{r}
head(all_trips_df)
```

### 4 - Analyze
#### 4.1 - Format columns
Upon further inspection, I realized that I needed to change the data types of the columns for later analysis.

```{r}
all_trips_df <- all_trips_df %>% 
  mutate(started_at = ymd_hms(started_at), # formats date as ymd hms
         ended_at = ymd_hms(ended_at), 
         start_station_id = as.numeric(start_station_id),
         end_station_id = as.numeric(end_station_id),
         start_lat = as.numeric(start_lat),
         start_lng = as.numeric(start_lng),
         end_lat = as.numeric(end_lat),
         end_lng = as.numeric(end_lng),
         ride_length_min = round(as.numeric(ended_at-started_at), digits = 2), # returns the diff in time in sec
         day_of_week = weekdays(started_at)) # retrieves day of the week that the ride started on

# change ride_length_min to min instead of sec
all_trips_df <- all_trips_df %>% 
  mutate(ride_length_min = round(as.numeric((ended_at-started_at)/60), digits = 2)) # returns diff in minutes

head(all_trips_df)
```


### 4.2 - Filter Data
```{r}
all_trips_df <- all_trips_df %>% 
  filter(start_station_name != "WATSON TESTING - DIVVY") # seemed to be a test location for the company, decided to filter out

all_trips_df <- all_trips_df %>% 
  filter(ride_length_min > 0) # filter out rides less than 0 minutes
```

### 4.3 - Mutate Data

#### Create df for Casual Riders

```{r}
casual_trips_df <- all_trips_df %>% # data frame of rides for casual riders
  filter(member_casual == "casual",
         ride_length_min < 1500, # filter out rides longer than 24hrs
         year(started_at) > "2020") %>% # filter out 2020 data
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), # create a hierarchy of weekdays
         month = month(started_at)) # create column representing the month the ride was in
```


#### Create df for Annual Members

```{r}
annual_member_trips_df <- all_trips_df %>% # data frame of rides for annual members
  filter(member_casual == "member", 
         ride_length_min < 1500, # filter out rides longer than 24hrs
         year(started_at) > "2020") %>% # filter out 2020 data
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), # create a hierarchy of weekdays
         month = month(started_at)) # create column representing the month the ride was in
```


### 4.4 - Summary Statistics
```{r}
summary(casual_trips_df)
```
```{r}
min(casual_trips_df$ride_length_min)
mean(casual_trips_df$ride_length_min)
max(casual_trips_df$ride_length_min)
```


```{r}
summary(annual_member_trips_df)
```
```{r}
min(annual_member_trips_df$ride_length_min)
mean(annual_member_trips_df$ride_length_min)
max(annual_member_trips_df$ride_length_min)
```

After a quick look at the summary statistics, we can see that casual riders spend more time on average per each ride than annual members. Casual riders spend and average of `r mean(casual_trips_df$ride_length_min)` minutes per ride, whereas annual members spend an average of `r mean(annual_member_trips_df$ride_length_min)` minutes.

## 5 - Plots

### 5.1 - Bar Charts
#### Casual Riders 
```{r cache=TRUE}
casual_weekly_plot <- ggplot(data = casual_trips_df, mapping = aes(x=day_of_week)) +
  geom_bar(aes(fill = day_of_week)) + # color for each weekday
  labs(
    title = "Number of Weekly Casual Rides",
    subtitle = "January 2021 - October 2021",
    x = "Day of the Week",
    y = "Number of Rides",
    fill = "Day of the Week"
  ) +
  theme(axis.text.x = element_text(angle = 90)) + # turns xaxis labels 90 degrees counterclockwise
  facet_wrap(~month) # create a weekly plot for each month
casual_weekly_plot
```



#### Annual Members
```{r cache=TRUE}
member_weekly_plot <- ggplot(data = annual_member_trips_df, mapping = aes(x=day_of_week)) +
  geom_bar(aes(fill = day_of_week)) + # color for each weekday
  labs(
    title = "Number of Weekly Member Rides",
    subtitle = "January 2021 - October 2021",
    x = "Day of the Week",
    y = "Number of Rides",
    fill = "Day of the Week"
  ) +
  theme(axis.text.x = element_text(angle = 90)) + # turns xaxis labels 90 degrees counterclockwise
  facet_wrap(~month) # create a weekly plot for each month
member_weekly_plot
```



### 5.2 - Density Plots

#### Map Setup
```{r map_set_up, message=FALSE, cache=TRUE}
###
## create map of chicago from https://cfss.uchicago.edu/notes/raster-maps-with-ggmap/
#
chi_bb <- c( # store bounding box coordinates
  left = -87.936287,
  bottom = 41.679835,
  right = -87.447052,
  top = 42.000835
)

chicago_stamen <- get_stamenmap( # retrieve map from Stamen Maps
  bbox = chi_bb,
  zoom = 11
)

chicago_map <- chicago_stamen
ggmap(chicago_map) # use ggmap to view 
```


#### Casual Riders
```{r}
casual_ride_density <- ggmap(chicago_map) + # map to show yearly ride density
  stat_density2d(data = casual_trips_df, aes(x=start_lng, y=start_lat), color="Deep Pink", fill=stat(nrow(casual_trips_df)), alpha=I(0.5), geom = "polygon") +
  labs(
    title = "Ride Density for Casual Riders",
    subtitle = "January 2021 - October 2021",
    x = "Longitude",
    y = "Latitude",
  )
```
```{r casual_ride_density_plot, cache=TRUE}
casual_ride_density
```


#### Annual Members
```{r}
member_ride_density <- ggmap(chicago_map) + # map to show ride yearly ride density
  stat_density2d(data = annual_member_trips_df, aes(x=start_lng, y=start_lat), color="Deep Pink", fill=stat(nrow(annual_member_trips_df)), alpha=I(0.5), geom = "polygon") +
  labs(
    title = "Ride Density for Members",
    subtitle = "January 2021 - October 2021",
    x = "Longitude",
    y = "Latitude",
  )
```
```{r member_ride_density_plot, cache=TRUE}
member_ride_density
```


## 6 - Conclusion
To sum up everthing that has been shown so far, we loaded the necessary packages and data, cleaned and filtered the data, manipulated the data frame to fit our needs, analyzed the data, and built some beautiful plots. Now it's time to circle back to our business problem. We had two questions to conquerer.

1. How do annual members and casual riders use Cyclistic bikes differently?
2. How can Cyclystic convert casual riders into annual members and boost profits?

### Findings
We found that for both casual and annual riders, the winter months (November, December, January, February, and most of March) are exceedingly lower in terms of the number of rides. And this makes sense, the winter months are typically cold, rainy, windy, and/or snowy in Chicago. Those conditions are not favorable for biking, even for the most dedicated users. However, the Spring, Summer, and early Fall months are the peak of Cyclystic's users' activity. The weather is fair, if not beautiful, allowing for an optimal environment for biking. 

We also found that there are differences in how casual and annual members use this service. Casual riders have high activity on the weekend days in comparison the weekdays. Annual members have a different spread of activity. These users utilize the bikes more equally throughout the week as a whole with a slight emphasis on weekdays rather than weekends. This could indicate that these users are using this service to commute to and from work or to run to get lunch during a busy day. 

Another interesting find was that most rides for both casual and annual riders where located in downtown Chicago. In addition, a lot of rides start up and down the water front area. We could infer that this is due to the city's contract with Cyclystic in terms of where bikes are allowed to be located to avoid having them littering the streets. We could also infer that it would be an optimal area for users to get around and see what the city has to offer in a more efficient way that walking or Ubering everywhere.

### Next Steps
Moving forward, there are some things I think Cyclystic should do if they want to convert casual riders into annual members.

1. Direct marketing efforts toward engaging users during the winter months.
2. Encourage more community building events (i.e. group rides) during weekdays to engage those who have historically only ridden on the weekends.
3. Expand geographical limits of the fleet to include more neighborhoods in the surrounding downtown area.


